---
title: "Second Markdown Doc"
author: "Keith Lewis"
date: "November 16, 2017"
output:
  pdf_document:
    fig_caption: yes
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(kfigr)
library(xtable)
library(knitr)

#file.exists("../data/trawl_abiotic.csv")
#file.exists("../data/trawl_biomass.csv")

# tangle the file
#purl(exercise_adv.Rmd, documentation = 2)

# load the data
abiotic <- read_csv("../data/trawl_abiotic.csv", guess_max = 5000)
head(abiotic)

# modify the data
abiotic_abrev <- filter(abiotic, year < 2005)
```

## Introduction
In a hole in the ground there lived a hobbit.....(`r figr('histo1', TRUE, type="Fig.")`)

```{r 'histo1', fig.cap="A histogram of bottom temperatures in NAFO division 2J.", echo=F, warning=FALSE, message=F}
abiotic_abrev %>%                        # Start with the 'abiotic' dataset
  filter(nafo_div == "2J") %>%     # Then, filter down to rows where nafo_div == 2J
  ggplot(aes(temp_bottom)) +       # Then, determine aes
  geom_histogram()               # plot histograms
```

Not a nasty dirty, wet hole filled with the ends of worms and an oozy smell, nor yet a dry, bare sandy hole with nothing in it to sit down on or to eat: it was a hobbit-hole, and that means comfort.

```{r echo = FALSE, results='asis'}
tab1 <- xtable(abiotic_abrev %>%
  group_by(nafo_div) %>%
  summarise(mean_depth = mean(depth), std_depth = sd(depth)))
print(tab1, comment = FALSE)
```

Across NAFO divisions, the grand mean of the depth was `r round(mean(abiotic_abrev$depth), 2)` (SD = `r round(sd(abiotic_abrev$depth), 2)`)
