---
title: "Project management: How to effectively organize a project in R"
author: "Paul Regular and Keith Lewis"
date: "NAFC | Fisheries and Oceans Canada | October 26, 2017"
output: 
  ioslides_presentation:
    incremental: true
    widescreen: true
    logo: rproj.png
---


```{r setup, echo=FALSE, results="hide", message=FALSE, warning=FALSE}
knitr::opts_knit$set(root.dir = '../')
knitr::opts_chunk$set(cache = TRUE, 
                      fig.align = "center", 
                      fig.height = 4.5, 
                      fig.width = 7.5,
                      dev = "svg")
```

## Examples of what not to do!!!
- uncommented code; write for yourself 6 mo. from now
- long chunks of code
- all files in one folder
- don't use functions
- lots of versions of a file
- no metadata
- no documentation of file/folder structure
- cut/paste results into MS WORD
Note: may want to add screenshots of above
## Best practices
- uncommented code; write for yourself 6 mo. from now
- long chunks of code; split code into tasks
- all files in one folder; folder for similar file types
- don't use functions; use'em
- lots of versions of a file; version control
- no metadata; make metadata
- no documentation of file/folder structure; README files
- cut/paste results into MS WORD; use markdown

## Outline of next 4 weeks
use figure

## Objectives {.build}

- Introduce the concept of a Research Compendium
- Demonstrate the set-up and workflow of a compendium
- Give a brief teaser of git and Rmarkdown

## Research Compendium

- What is it?
    - A compendium is a convention for how to organize key files for a project (data, code, text, etc.) into directories
- Why use one?
    - Simplifies file management
    - Streamlines analytical workflows
    - More efficient for you, future you and your collaborators
    - Reproducible and extensible
- How should files be organized?
    - Apply the conventions used for building an R package
    
## Getting started {.smaller}

Use the `devtools` package to set-up the structure

```{r, eval=FALSE}

## Project path and name
root_dir <- "examples"                                  # path to your project
proj_name <- "skeleton"                                 # project name

## Metadata for the project
proj_metadata <- list("Title" = "Project title",
                      "Authors@R" = person("First", "Last", 
                                          email = "first.last@example.com", 
                                          role = c("aut", "cre")),
                      "Description" = "This is a research compendium for...",
                      "License" = "NA")

## Generate the skeleton for the project
devtools::create(file.path(root_dir, proj_name),
                 description = proj_metadata)           # generates a minimum R package skeleton
devtools::use_git(pkg = file.path(root_dir, proj_name)) # Initalize local git repository
dir.create(file.path(root_dir, proj_name, "data"))      # adds a data folder to the directory
dir.create(file.path(root_dir, proj_name, "analysis"))  # adds an analysis folder to the directory

```


## Core structure {.smaller}

Code on the last slide generates a structure similar to the one reccomended here: https://github.com/ropensci/rrrpkg

```
skeleton
|
|- skeleton.Rproj       # RStudio Rproj file; one file to rule them all
|
|- DESCRIPTION          # project metadata and dependencies 
|
|- data/                # raw data, not changed once created
|  +- my_data.csv       # data files in open formats such as TXT, CSV, TSV, etc.
|
|- R/                   # reusable code
|  +- my_functions.R    # custom R functions that are used more than once in the project
|
|- analysis/            # any programmatic code 
|  +- my_scripts.R      # R code or R markdown files used to analyse and visualise data 

```

## Workflow - step 1 {.smaller}

Setup the compendium

```{r, eval=FALSE}

## Project path and name
root_dir <- "examples"                                  # path to your project
proj_name <- "strat"                                    # project name

## Metadata for the project
proj_metadata <- list("Title" = "Stratified analysis of shrimp, cod, halibut and redfish population trends",
                      "Authors@R" = c(person("Paul", "Regular", 
                                             email = "Paul.Regular@dfo-mpo.gc.ca", 
                                             role = c("aut", "cre")),
                                      person("Keith", "Lewis", 
                                             email = "Keith.Lewis@dfo-mpo.gc.ca", 
                                             role = "aut")),
                                      "Description" = "This is a research compendium for stratified
                                                       analysis of shrimp population trends.",
                                      "License" = "NA")

## Generate the skeleton for the project
devtools::create(file.path(root_dir, proj_name),
                 description = proj_metadata)           # generates a minimum R package skeleton
devtools::use_git(pkg = file.path(root_dir, proj_name)) # Initalize local git repository
dir.create(file.path(root_dir, proj_name, "data"))      # adds a data folder to the directory
dir.create(file.path(root_dir, proj_name, "analysis"))  # adds an analysis folder to the directory

```

## Workflow - step 2 {.build}

Add the data

```
|- data/
|-  +- trawl_abiotic.csv
|-  +- trawl_biotic.csv

```
- These data should ramin static


## Workflow - step 3

Open `strat.Rproj` in RStudio
- Workinig directory will automatically = project directory

## Workflow - step 4 {.build}

Save an R scritp in the analysis folder and start writing code to manipulate, visualize and model the data
```
|- analysis/
|-   +- strat_analysis.R
```
- Work from the project directory (e.g. `read.csv("data/trawl_biotic.csv")`)
- If your script grows too long, it may be useful to split and number scripts
    ```
    |- analysis/
    |-   +- 001_manipulate.R
    |-   +- 002_visualize.R
    |-   +- 003_model.R
    ```

## Workflow - conditional steps

- Functions (+ taking advantage of an R package setup)
- Rmarkdown
- git
- TODO: add teaser slides on these topics



## Exercises
- Create a project folder and an RProj
- Add subfolders
- Write a script to:
  - subset some data
  - make some figures
  - save to a folder
  

## References

> - This presentation was heavily based on this really useful doc: https://github.com/ropensci/rrrpkg
> - For more details on building R packages, see: [Wickham, Hadley, R Packages: Organise, test, document, and share your code. O'Reilly.](http://r-pkgs.had.co.nz/)
> - Keep an eye on in development packages to simplify reproducable research in R: https://github.com/benmarwick/rrtools

