---
title: "Introduction to R"
subtitle: "Outline and set-up"
author: "Paul Regular, Eric Pedersen & Aaron Adamack"
date: "NAFC | Fisheries and Oceans Canada | 2019-01-16"
output: 
  ioslides_presentation:
    logo: Rlogo.png
    incremental: true
    widescreen: true
---



## Outline

| Time           | Topic                                                                    |
| :------------- | ------------------------------------------------------------------------ |
| 10:00 – 10:30  | Get Rstudio and R packages up and running, and set-up an R project       |
| 10:30 – 11:00  | Some basics of R coding [objects, functions, common operations]          |
| 11:00 – 12:00  | Import data [introduce data types and concept of tidy data]              |
| 12:00 – 13:00  | Lunch                                                                    |
| 14:00 – 15:00  | Visualize data [plots and facet plots]                                   |
| 13:00 – 14:00  | Transform data [filter, mutate, join and summarize data]                 |


## Set-up

- Do you have R and Rstudio installed?
- Are you happy with the layout of Rstudio?
- Have you turned off the save working directory prompt?


## Outcome

>- We hope you'll be able to write code to explore your data using R
>- A quick example of the work flow is shown in the next few slides

```{r message = FALSE, warning = FALSE}

## Packages used for this example
library(readxl)  # for reading excel data
library(tidyr)   # for data tidying functions
library(dplyr)   # for data manipulation functions
library(ggplot2) # for plotting

```

## Import {.smaller}

```{r}

## Import: get excel data into R
duck <- read_excel("data/Duck Islands Movement.xlsx")
round <- read_excel("data/Round Island Movement.xlsx")
round

```

## Tidy {.smaller}

```{r}

## Tidy: combine and clean-up these data
tags <- bind_rows(duck, round) %>% 
    select(SIDE:NOTCH) %>%         
    drop_na()  %>%                
    rename_all(tolower) 
tags

```

## Explore {.smaller}

```{r, fig.width = 10, fig.height = 4}

## Visualize: take an initial quick look at the data
ggplot(tags, aes(x = year, y = length, group = tag, color = side)) +
    geom_line() + facet_grid(sex ~ island)

```

## Explore {.smaller}

```{r}

## Transform: calculate growth rates of each tagged individual
rates <- tags %>% 
    group_by(tag, sex, island) %>% 
    arrange(tag, year) %>% 
    summarise(first_year = first(year), 
              last_year = last(year),
              first_length = first(length), 
              last_length = last(length)) %>% 
    filter(last_year > first_year, last_length > first_length) %>% 
    mutate(growth_rate = (last_year - first_year) / (last_length - first_length))
head(rates, 5)

```

## Explore {.smaller}

```{r, fig.width = 10, fig.height = 4}

## Visualize: plot raw growth rates
ggplot(rates, aes(x = sex, y = growth_rate, color = island)) +
    geom_jitter()

```

## Explore {.smaller}

```{r}

## Transform: calculate mean growth rates
mean_rates <- rates %>% 
    group_by(sex, island) %>% 
    summarise(mean = mean(growth_rate), sd = sd(growth_rate), 
              se = sd(growth_rate) / sqrt(n()), n = n())
mean_rates

```

## Explore {.smaller}

```{r}

## Visualize: produce a publication quality plot
dodge <- position_dodge(width = 0.9)
p <- ggplot(mean_rates, aes(x = sex, y = mean, fill = island)) + 
    geom_col(position = dodge, color = "black") +
    geom_errorbar(aes(ymin = mean - 2 * se, ymax = mean + 2 * se),
                  position = dodge, width = 0.1) +
    scale_fill_manual(values = c("grey50", "grey75"), 
                      name = "Island", labels = c("Duck", "Round")) +
    scale_y_continuous(expand = c(0, 0)) +
    scale_x_discrete(labels = c("Female", "Male")) +
    xlab("") + ylab("Mean growth rate (cm / year)") +
    theme_classic()

```


## Explore {.smaller}

```{r, fig.width = 7, fig.height = 4.5}
p # print plot
```

