---
title: "Introduction to R"
subtitle: "The basics"
author: "Paul Regular & Alejandro Buren"
date: "NAFC | Fisheries and Oceans Canada | 2019-01-16"
output: 
  ioslides_presentation:
    logo: Rlogo.png
    incremental: true
    widescreen: true
---


## What is R?

- A language and environment for statistical computing and graphics
- Free, open source, and available on Windows, Mac and Linux
- Flexible, expansive and cutting-edge
- While the learning curve is steep, there is a great community of users willing to help

## R Basics

- R holds **functions** and **data** in named **objects**
- Objects may be created by the user or loaded from a **package**


## Objects

- R has symbolic objects which can be assigned values
- The traditional way to do this in R is the `<-` operator
- Possible vector names flexible
- Vector names cannot start with a digit
- Names are case-sensitive
- Some common names are already used by R, so it's best not to use these
    - `c, q, t, C, D, F, I, T`

## Functions

- Functions are objects that do something
- There are many built in functions in `base` R
    - `log(), exp(), sin(), sqrt(), mean(), max(), ...`
- There are help files for most functions
    - Access help by entering `?fun_name`

## Data

- Data are also held in objects
- There are many ways to structure data
    - Vectors (numeric, character, logical, factor)
    - Matrix
        - All columns in a matrix must have same mode
    - Array
        - N-dimensional matrix
    - Data frame
        - Columns in a data frame can have different modes
    - Lists
        - Collection of objects, can be different modes and dimensions

## Packages

- Collections of R functions, data, and compiled code
- Custom built to for solving specific problems
- R comes with a standard set of packages
- Thousands are available on the Comprehensive R Archive Network (CRAN)


## Basic exercise {.smaller}

Use R as a calculator

```{r, eval = FALSE}
sqrt(80 / 20) * 3 # use R as a calculator
```

Create an object

```{r, eval = FALSE}
my_object <- sqrt(80 / 20)
my_object * 3
```

Why doesn't this work?

```{r, eval = FALSE}
x <- 17
X * 20
```

`## Error: object 'X' not found`

