---
title: "Introduction to R"
subtitle: "Plotting in the tidyverse: ggplot2"
author: "Paul Regular & Keith Lewis"
date: "NAFC | Fisheries and Oceans Canada | 2019-01-16"
output:
  ioslides_presentation:
    logo: ggplot2_logo.png
    incremental: yes
    widescreen: yes
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(fig.align = "center", 
                      fig.height = 4.5, 
                      fig.width = 8,
                      dev = "svg",
                      warning = FALSE,
                      message = FALSE)
```



## Outline

>- Plotting options
>- Grammar of graphics
>- ggplot2: the basics
>- ggplot2: intermediate
>- Summary


## Plotting options

*"The greatest value of a picture is when it forces us to notice what we never expected to see."* — John Tukey

- Spreadsheets (e.g., Excel)
- Stand alone program (e.g., Sigma Plot)
- Other stats programs (e.g., SAS)
- R
    - Base
    - Various functions
    - ggplot2 (part of the tidyverse)

## Grammar of graphics

- ggplot2 is based on the **grammar of graphics**: the idea that you can build every graph from a few components:
   - **Data**
       - Variables mapped to aesthetic properties (e.g. x, y, size, colour)
   - **Geometry**
       - Visual markers (e.g. points, lines, polygons)
   - **Coordinates**
       - A coordinate system (e.g. Cartesian, polar)

## ggplot2: the basics

>- Data
>- Syntax
>- Scatterplot
>- Line plot
>- Boxplot
>- Stacked bar
>- Exercise

## ggplot2: the basics - data {.build}

>- ggplot2 is built for tidy data
>- Below we import and tidy up lobster movement data

```{r}

library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)

duck <- read_excel("data/Duck Islands Movement.xlsx")
round <- read_excel("data/Round Island Movement.xlsx")
tags <- bind_rows(duck, round) %>%   # stack both data sets
    select(SIDE:NOTCH) %>%           # select main columns
    drop_na(SIDE, ISLAND, YEAR, TAG) # drop empty rows

```


## ggplot2: the basics - syntax {.build .smaller}

```{r, eval = FALSE}
ggplot(data = tags, mapping = aes(x = YEAR, y = LENGTH)) + # Data
    geom_point() +                                         # Geometry
    coord_cartesian()                                      # Coordinates
```

- **Data**
    - Here the `ggplot` function starts the plot
        - `data` is supplied first
        - `mapping` is specified second using the `aes` (aesthetics) function
- **Geometry**
    - The function `geom_point` adds points to the base layer
- **Goordinates**
    - Finally, `coord_cartesian` defines the coordinate system
        - This is the default so it is not usually specified

> *Note that the pipe operator (`%>%`) is not used here, rather you add items (`+`)*
    

## ggplot2: the basics - scatterplot  {.build}

```{r}
ggplot(tags, aes(x = YEAR, y = LENGTH)) + 
    geom_point()
```

## ggplot2: the basics - line plot  {.build}

```{r}
ggplot(tags, aes(x = YEAR, y = LENGTH, group = TAG)) + 
    geom_line()
```

## ggplot2: the basics - box-whisker plot  {.build}

```{r}
ggplot(tags, aes(x = SEX, y = LENGTH)) + 
    geom_boxplot()
```

## ggplot2: the basics - stacked bargraph  {.build}

```{r}
ggplot(tags, aes(x = YEAR, fill = ISLAND)) +
  geom_bar() 
```

## ggplot2: the basics - exercises 

```{r, eval = FALSE}
ggplot(tags, aes(x = YEAR, y = LENGTH, group = TAG)) + 
    geom_line()
```

Add the `colour` aesthetic to the above code make the plot below

```{r, echo = FALSE, fig.height = 3.5}
ggplot(tags, aes(x = YEAR, y = LENGTH, group = TAG, colour = SEX)) + 
    geom_line()
```


## ggplot2: intermediate

Move away from defaults to publication quality

>- Build upon an object
>- Change size and colour
>- Facets
>- Axes
>- Legends
>- Themes

## ggplot2: intermediate - object + geom {.build}

```{r}
p <- ggplot(tags)
p + geom_line(aes(x = YEAR, y = LENGTH, group = TAG, colour = SEX))
```

## ggplot2: intermediate - object + alternate aes {.build}

```{r}
p + geom_line(aes(x = YEAR, y = LENGTH, group = TAG, color = ISLAND))
```

## ggplot2: intermediate - object + alternate aes {.build}

```{r}
p + geom_line(aes(x = YEAR, y = LENGTH, group = TAG, color = SIDE))
```


## ggplot2: intermediate - object + alternate geom {.build}

```{r}
p + geom_bar(aes(x = YEAR, fill = ISLAND), stat = "count")
```


## ggplot2: intermediate - object + geom + facet {.smaller .build}

```{r}
p + geom_bar(aes(x = YEAR)) +
    facet_grid(~ ISLAND) # split plots by island
```

## ggplot2: intermediate - object + geom + facet {.smaller .build}

```{r}
p + geom_bar(aes(x = YEAR)) +
    facet_grid(SEX ~ ISLAND) # split plots by sex and island
```

## ggplot2: intermediate - object + geom + facet {.smaller .build}

```{r}
p + geom_bar(aes(x = YEAR, fill = factor(MONTH))) +
    facet_grid(SEX ~ ISLAND)
```

## ggplot2: intermediate - object + more grammar {.build}

- Let's move away from the defaults by iteratively modifying the plot in the previous slide 

```{r}
## Save the first step into an object called p1
p1 <- ggplot(tags) + 
    geom_bar(aes(x = YEAR, fill = factor(MONTH))) +
    facet_grid(SEX ~ ISLAND)
```


## ggplot2: intermediate - object + more grammar {.smaller}

```{r}
## Here's what the defaults look like
p1
```


## ggplot2: intermediate - object + more grammar {.smaller}

```{r}
p2 <- p1 + xlab("Year") + ylab("Number of records") # change x and y labels
p2
```


## ggplot2: intermediate - object + more grammar {.smaller}

```{r}
p3 <- p2 + scale_fill_brewer(palette = "RdBu", name = "Month")  # name legend and use greyscale
p3
```

## ggplot2: intermediate - object + more grammar {.smaller}

```{r}
p4 <- p3 + coord_cartesian(ylim = c(0, 700), expand = FALSE) # set y limits and remove buffer
p4
```

## ggplot2: intermediate - object + more grammar {.smaller}

```{r}
p5 <- p4 + theme_bw() # use the black and white theme
p5
```

## ggplot2: intermediate - object + more grammar {.smaller}

```{r}
p6 <- p5 + theme(panel.grid.major = element_blank()) # remove major grid lines
p6
```

## ggplot2: intermediate - object + more grammar {.smaller}

```{r}
p7 <- p6 + theme(panel.grid.minor = element_blank()) # remove minor grid lines
p7
```

## ggplot2: intermediate - object + more grammar {.smaller}

```{r}
p8 <- p7 + theme(strip.background = element_blank()) # remove facet strip background
p8
```


## ggplot2: intermediate - object + more grammar {.smaller}

>- All together plus an option for changing the facet labels

```{r}

## Set-up nice island and sex labels
island_names <- c("duck" = "Duck",
                  "round" = "Round")
sex_names <- c("F" = "Female",
               "M" = "Male")

## Store the plot in an object
p <- ggplot(tags) +                                                 # set-up base layer
    geom_bar(aes(x = YEAR, fill = factor(MONTH)), stat = "count") + # add bars by year and filled by month
    facet_grid(SEX ~ ISLAND,                                        # facet by sex and island
               labeller = labeller(SEX = sex_names,                 # edit facet labels
                                   ISLAND = island_names)) +        
    xlab("Year") + ylab("Number of records") +                      # change x and y labels
    scale_fill_brewer(palette = "RdBu", name = "Month") +         # name legend and use greyscale
    coord_cartesian(ylim = c(0, 700), expand = FALSE) +             # set y limits and remove buffer
    theme_bw() +                                                    # use the black and white theme
    theme(panel.grid.major = element_blank(),                       # remove major grid lines
          panel.grid.minor = element_blank(),                       # remove minor grid lines
          strip.background = element_blank())                       # remove facet strip background

```

## ggplot2: intermediate - object + more grammar {.smaller}

```{r}
p
```

## Summary

- ggplot2 is a very flexible visualization tool
    - Have only shown a small subset of what it can do
- Plots can help you discover patterns in your data
    - They can also uncover issues in the data
- Your ability to explore your data will expand when you pair `ggplot2` and `dplyr`





