---
title: "Introduction to R"
subtitle: "Working with tidy data: mutating, grouping, and summarizing"
author: "Eric Pedersen, Keith Lewis & Paul Regular"
date: "NAFC | Fisheries and Oceans Canada | 2019-01-16"
output: 
  ioslides_presentation:
    logo: dplyr_logo.png
    incremental: true
    widescreen: true
---

## What we hope you'll be able to do by the end of this presentation:

```{r setup_run, echo = FALSE, eval = TRUE,include=FALSE}
library(knitr)
library(readxl)
library(tidyr)
library(dplyr)
library(ggplot2) 
opts_chunk$set(cache = TRUE, autodep = TRUE)

duck <- read_excel("data/Duck Islands Movement.xlsx")
round <- read_excel("data/Round Island Movement.xlsx")

tags <- bind_rows(duck, round)
tags

tidy_tags <- select(tags, SIDE:NOTCH) %>% 
    drop_na(SIDE, ISLAND, YEAR, TAG)%>%
    filter(SIDE!="s")
```

```{r length_full_plot, echo = FALSE, eval = TRUE ,purl=FALSE}
legal_tags <- tidy_tags %>%
    mutate(LEGAL_SIZE = LENGTH > 82.5,
           SIDE = tolower(SIDE)) %>%
    filter(LEGAL_SIZE == TRUE) %>%
    group_by(TAG,SEX, ISLAND)%>%
    mutate(YEARS_OBS = max(YEAR)-min(YEAR))%>%
    filter(YEARS_OBS >0)%>%
    ungroup()

longest_observed_lobsters <- legal_tags %>%
    group_by(SEX,ISLAND)%>%
    filter(YEARS_OBS == max(YEARS_OBS))%>%
    ungroup()

size_plot <- legal_tags %>%
    ggplot(aes(x = YEAR, y = LENGTH, color = SEX))+
    facet_grid(SEX~ISLAND)+
    geom_line(aes(group = TAG),alpha=0.25)+
    geom_line(data= longest_observed_lobsters, 
              aes(group = TAG),
              size=2)+
    scale_color_brewer(palette = "Set1")+
    theme_bw()+
    theme(panel.grid = element_blank())


print(size_plot)

```

## What we hope you'll be able to do by the end of this presentation:

```{r growth_full_plot, echo = FALSE, eval = TRUE,purl=FALSE}



growth_rate_summary = legal_tags %>%
    group_by(TAG, SEX, ISLAND)%>%
    filter(max(YEAR)>min(YEAR))%>%
    summarize(LENGTH_START = min(LENGTH),
              GROWTH_RATE = (max(LENGTH)-min(LENGTH))/(max(YEAR)-min(YEAR)))%>%
    ungroup()


growth_plot = growth_rate_summary %>%
    ggplot(aes(x = LENGTH_START, GROWTH_RATE,color = SEX))+
    facet_grid(SEX~ISLAND)+
    geom_point()+
    scale_color_brewer(palette = "Set1")+
    theme_bw()+
    theme(panel.grid = element_blank())

print(growth_plot)

```

## dplyr is:

* A tool for working with tidy data
* Works by using simple verbs, that each do one thing well: 
    * `mutate(data, <other arguments>)`
    * `arrange(data, <other arguments>)`
    * `group_by(data, <other arguments>)`
    * `summarize(data, <other arguments>)`
    * ...
* Lets you apply these verbs in order to create new tidy dataframes:

* `new_data <- old_data %>%
     select(<something>)%>%
     mutate(<something else>)%>%
     ...`

## First make sure we're on the same page:

```{r setup, echo = TRUE, eval=FALSE}
<<setup_run>>
```

## `mutate()`: Creating and altering exisiting variables

* `mutate()` lets you change columns, or create new columns
* You can add as many columns as you want in one `mutate` statement
* it does not add or remove rows, change the order of the data, or not remove columns

## `mutate()`: Creating and altering exisiting variables 

```{r mutate_demo, eval = FALSE, echo=TRUE, purl=TRUE}
#Let's say we want to see how catches of legal (82.5 mm+) and non-legal size 
#have changed over time in the tagging data, compared to the average size of all
#lobsters caught:
tidy_tags <- tidy_tags %>%
    mutate(LEGAL_SIZE = LENGTH > 82.5,
           LENGTH_AVERAGE = mean(LENGTH),
           #We can also change an existing variable: 
           SIDE = tolower(SIDE)
           )

tidy_tags
```

## `mutate()`: Creating and altering exisiting variables 

```{r mutate_demo_eval, eval = TRUE, echo=FALSE, purl=FALSE}
<<mutate_demo>>
```

## `mutate()`: Creating and altering exisiting variables 

```{r mutate_demo2, eval = FALSE, echo=TRUE, purl=TRUE}
#Now we can plot this:
legal_size_plot <- ggplot(tidy_tags, aes(x=YEAR, y= LENGTH, color = LEGAL_SIZE))+
    facet_grid(ISLAND~SIDE)+
    geom_point()+
    geom_hline(aes(yintercept = LENGTH_AVERAGE))

print(legal_size_plot)
```

## `mutate()`: Creating and altering exisiting variables 

```{r mutate_demo_plot, eval = TRUE, echo=FALSE}
<<mutate_demo2>>
```

## Exercise: working with tidyr

Create a new data frame called `legal_size_tags`, that takes the `tidy_tags` data frame, creates the new `LEGAL_SIZE` column, and then use the `filter` command to exclude all of the lobsters than are less than legal size.

```{r mutate_exercise, eval = TRUE, echo=FALSE}
#Exercise 1####

#Create a new data frame called `legal_size_tags`, that takes the `tidy_tags`
#data frame, creates the new `LEGAL_SIZE` column, and then use the `filter`
#command to exclude all of the lobsters than are less than legal size.
```

## Exercise: working with tidyr

Create a new data frame called `legal_size_tags`, that takes the `tidy_tags` data frame, creates the new `LEGAL_SIZE` column, and then use the `filter` command to exclude all of the lobsters than are less than legal size.
```{r mutate_exercise_results, eval = TRUE, echo=TRUE, purl = FALSE}
legal_tags <- tidy_tags %>%
    mutate(LEGAL_SIZE = LENGTH > 82.5,
           SIDE = tolower(SIDE)) %>%
    filter(LEGAL_SIZE == TRUE) 

legal_tags
```








## To do for tomorrow:
* `group_by`
* `summarize`
