---
title: "Introduction to R"
subtitle: "Working with tidy data: mutating, grouping, and summarizing"
author: "Eric Pedersen, Keith Lewis & Paul Regular"
date: "NAFC | Fisheries and Oceans Canada | 2019-01-16"
output: 
  ioslides_presentation:
    logo: dplyr_logo.png
    incremental: true
    widescreen: true
---

## What we hope you'll be able to do by the end of this presentation:

```{r knitr_options, echo = FALSE, eval = TRUE, include = FALSE, purl = FALSE}
library(knitr)
opts_chunk$set(cache = TRUE, autodep = TRUE)
```

```{r setup_run, echo = FALSE, eval = TRUE,include=FALSE, purl = FALSE}
library(readxl)
library(tidyr)
library(dplyr)
library(ggplot2) 


duck <- read_excel("data/Duck Islands Movement.xlsx")
round <- read_excel("data/Round Island Movement.xlsx")

tags <- bind_rows(duck, round)
tags

tidy_tags <- select(tags, SIDE:NOTCH) %>% 
    drop_na(SIDE, ISLAND, YEAR, TAG)%>%
    filter(SIDE!="s")
```

```{r length_full_plot, echo = FALSE, eval = TRUE ,purl=FALSE}
legal_tags <- tidy_tags %>%
    mutate(LEGAL_SIZE = LENGTH > 82.5,
           SIDE = tolower(SIDE)) %>%
    filter(LEGAL_SIZE == TRUE) %>%
    group_by(TAG,SEX, ISLAND)%>%
    mutate(YEARS_OBS = max(YEAR)-min(YEAR))%>%
    filter(YEARS_OBS >0)%>%
    ungroup()

longest_observed_lobsters <- legal_tags %>%
    group_by(SEX,ISLAND)%>%
    filter(YEARS_OBS == max(YEARS_OBS))%>%
    ungroup()

size_plot <- legal_tags %>%
    ggplot(aes(x = YEAR, y = LENGTH, color = SEX))+
    facet_grid(SEX~ISLAND)+
    geom_line(aes(group = TAG),alpha=0.25)+
    geom_line(data= longest_observed_lobsters, 
              aes(group = TAG),
              size=2)+
    scale_color_brewer(palette = "Set1")+
    theme_bw()+
    theme(panel.grid = element_blank())


print(size_plot)

```

## What we hope you'll be able to do by the end of this presentation:

```{r growth_full_plot, echo = FALSE, eval = TRUE,purl=FALSE}



lobster_growth_rates = legal_tags %>%
    group_by(TAG, SEX, ISLAND)%>%
    filter(max(YEAR)>min(YEAR))%>%
    summarize(LENGTH_START = min(LENGTH),
              GROWTH_RATE = (max(LENGTH)-min(LENGTH))/(max(YEAR)-min(YEAR)))%>%
    ungroup()


growth_plot = lobster_growth_rates %>%
    ggplot(aes(x = LENGTH_START, GROWTH_RATE,color = SEX))+
    facet_grid(SEX~ISLAND)+
    geom_point()+
    #This tweaks the colors into something nicer
    scale_color_brewer(palette = "Set1")+
    #These change the elements of the theme
    theme_bw()+
    theme(panel.grid = element_blank())

print(growth_plot)

```

## dplyr is:

* A tool for working with tidy data
* Works by using simple verbs, that each do one thing well: 
    * `mutate(data, <other arguments>)`
    * `arrange(data, <other arguments>)`
    * `group_by(data, <other arguments>)`
    * `summarize(data, <other arguments>)`
    * ...
* Lets you apply these verbs in order to create new tidy dataframes:

* `new_data <- old_data %>%
     select(<something>)%>%
     mutate(<something else>)%>%
     ...`

## First make sure we're on the same page:

```{r setup, echo = TRUE, eval=FALSE}
<<setup_run>>
```

## `mutate()`: Creating and altering exisiting variables

* `mutate()` lets you change columns, or create new columns
* You can add as many columns as you want in one `mutate` statement
* it does not add or remove rows, change the order of the data, or not remove columns

## `mutate()`: Creating and altering exisiting variables 

```{r mutate_demo, eval = FALSE, echo=TRUE, purl=TRUE}
#Let's say we want to see how catches of legal (82.5 mm+) and non-legal size 
#have changed over time in the tagging data, compared to the average size of all
#lobsters caught:
tidy_tags <- tidy_tags %>%
    mutate(LEGAL_SIZE = LENGTH > 82.5,
           LENGTH_AVERAGE = mean(LENGTH),
           #We can also change an existing variable: 
           SIDE = tolower(SIDE)
           )

tidy_tags
```

## `mutate()`: Creating and altering exisiting variables 

```{r mutate_demo_eval, eval = TRUE, echo=FALSE, purl=FALSE}
<<mutate_demo>>
```

## `mutate()`: Creating and altering exisiting variables 

```{r mutate_demo2, eval = FALSE, echo=TRUE, purl=TRUE}
#Now we can plot this:
legal_size_plot <- ggplot(tidy_tags, aes(x=YEAR, y= LENGTH, color = LEGAL_SIZE))+
    facet_grid(ISLAND~SIDE)+
    geom_point()+
    geom_hline(aes(yintercept = LENGTH_AVERAGE))

print(legal_size_plot)
```

## `mutate()`: Creating and altering exisiting variables 

```{r mutate_demo_plot, eval = TRUE, echo=FALSE}
<<mutate_demo2>>
```

## Exercise: working with `mutate`

Create a new data frame called `legal_size_tags`, that takes the `tidy_tags` data frame, creates the new `LEGAL_SIZE` column, and then use the `filter` command to exclude all of the lobsters than are less than legal size.

```{r mutate_exercise, eval = TRUE, echo=FALSE}
#Exercise 1####

#Create a new data frame called `legal_size_tags`, that takes the `tidy_tags`
#data frame, creates the new `LEGAL_SIZE` column, and then use the `filter`
#command to exclude all of the lobsters than are less than legal size.
```

## Exercise: working with `mutate`

Create a new data frame called `legal_size_tags`, that takes the `tidy_tags` data frame, creates the new `LEGAL_SIZE` column, and then use the `filter` command to exclude all of the lobsters than are less than legal size.
```{r mutate_exercise_results, eval = TRUE, echo=TRUE, purl = FALSE}
legal_tags <- tidy_tags %>%
    mutate(LEGAL_SIZE = LENGTH > 82.5,
           SIDE = tolower(SIDE)) %>%
    filter(LEGAL_SIZE == TRUE) 

legal_tags
```



## `group_by()`: breaking data into groups 

* `group_by()` is a function for telling `dplyr` that any further analyses should be done *within the selected groups*
* by itself, it does not change how the data look at all,
 but it affects all the stages after it
* Can group by a *single* variable, or *multiple* variables
* You can stop grouping by using `ungroup()`



## `group_by()`: breaking data into groups 

```{r groupby_intro, eval = FALSE, echo=TRUE, purl = TRUE}
tidy_tags <- tidy_tags %>%
    group_by(SEX)

tidy_tags
```

## `group_by()`: breaking data into groups 
```{r groupby_intro_eval, eval = TRUE, echo=TRUE, purl = FALSE}
<<groupby_intro>>
```

## `group_by()`: breaking data into groups 

```{r groupby_intro2, eval = FALSE, echo=TRUE, purl = TRUE}
#Here's how you group with multiple different variables
tidy_tags <- tidy_tags %>%
    group_by(SIDE, ISLAND)

tidy_tags
```

## `group_by()`: breaking data into groups 
```{r groupby_intro2_eval, eval = TRUE, echo=TRUE, purl = FALSE}
<<groupby_intro2>>
```

## `group_by()`: breaking data into groups 
`group_by() is not that useful alone, but is incredibly powerful when combined with other commands:
```{r groupby_plot, eval = FALSE, echo=TRUE, purl = TRUE}
tidy_tags <- tidy_tags %>%
    group_by(SIDE, ISLAND) %>%
    mutate(LENGTH_AVG = mean(LENGTH))%>%
    #It's always important to use ungroup after doing the grouped action
    ungroup()

#Now we can plot this:
legal_size_plot <- ggplot(tidy_tags, aes(x=YEAR, y= LENGTH, color = LEGAL_SIZE))+
    facet_grid(ISLAND~SIDE)+
    geom_point()+
    geom_hline(aes(yintercept = LENGTH_AVG))

print(legal_size_plot)
```

## `group_by()`: breaking data into groups 

```{r  groupby_plot_eval, eval = TRUE, echo=FALSE}
<<groupby_plot>>
```

## Exercise: working with `group_by`

Modify the following code so that you are calculating the length of time that each tag value has been observed for each sex and each island (remember to use ungroup at the end!)

```{r groupby_exercise_comments, eval = TRUE, echo=FALSE}
#Exercise 2####

# Modify the following code so that you are calculating the length of time that 
# each tag value has been observed for each sex and each island (remember to use ungroup at the end!)
```

```{r groupby_exercise, eval = FALSE, echo=TRUE, purl  = TRUE}
legal_tags <- tidy_tags %>%
    mutate(LEGAL_SIZE = LENGTH > 82.5,
           SIDE = tolower(SIDE)) %>%
    filter(LEGAL_SIZE == TRUE) %>%
    mutate(YEARS_OBS = max(YEAR)-min(YEAR))%>%
    filter(YEARS_OBS >0)
```


## Exercise: working with `group_by`

Modify the following code so that you are calculating the length of time that each tag value has been observed for each sex and each island (remember to use ungroup at the end!)

```{r groupby_exercise_results, eval = TRUE, echo=TRUE, purl = FALSE}
legal_tags <- tidy_tags %>%
    mutate(LEGAL_SIZE = LENGTH > 82.5,
           SIDE = tolower(SIDE)) %>%
    filter(LEGAL_SIZE == TRUE) %>%
    group_by(TAG,SEX, ISLAND)%>%
    mutate(YEARS_OBS = max(YEAR)-min(YEAR))%>%
    filter(YEARS_OBS >0)%>%
    ungroup()
```

## Now we can create the first plot we saw:

```{r groupby_data, eval = TRUE, echo=TRUE, purl = TRUE}
legal_tags <- tidy_tags %>%
    mutate(LEGAL_SIZE = LENGTH > 82.5,
           SIDE = tolower(SIDE)) %>%
    filter(LEGAL_SIZE == TRUE) %>%
    group_by(TAG,SEX, ISLAND)%>%
    mutate(YEARS_OBS = max(YEAR)-min(YEAR))%>%
    filter(YEARS_OBS >0)%>%
    ungroup()

longest_observed_lobsters <- legal_tags %>%
    group_by(SEX,ISLAND)%>%
    filter(YEARS_OBS == max(YEARS_OBS))%>%
    ungroup()
```



## Now we can (almost) create the first plot we saw:
```{r groupby_plot2, eval = FALSE, echo=TRUE, purl = TRUE}
size_plot <- legal_tags %>%
    ggplot(aes(x = YEAR, y = LENGTH, color = SEX))+
    facet_grid(SEX~ISLAND)+
    geom_line(aes(group = TAG),alpha=0.25)+
    geom_line(data= longest_observed_lobsters, 
              aes(group = TAG),
              size=2)

print(size_plot)
```

## Now we can almost create the first plot we saw:

```{r groupby_plot3, eval = TRUE, echo=FALSE, purl = FALSE}
<<groupby_plot2>>
```

(There's code at the end of the script file we sent you that shows how to make this figure even prettier)


